Isomorphisms as Groupoids: \\ Homotopy Type Theory with Fractional Types

### 2013-10-16 f.tex
paths

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{color, colortbl}
\usepackage{courier}
\usepackage{graphicx}

@@

\newcommand{\union}{\cup}
\newcommand{\transport}[2]{\overline{#1}(#2)}
\newtheorem{definition}{Definition}
\newcommand{\refl}{\mathit{refl}}
\renewcommand{\path}{\leadsto}
%% \renewcommand{\circ}{\circlearrowleft}
\newcommand{\alt}{~|~}
\newcommand{\dt}[1]{\llbracket #1 \rrbracket}
\newcommand{\leftv}[1]{\textsf{left}~#1}
\newcommand{\rightv}[1]{\textsf{right}~#1}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\wedgesum}{\vee}
\newcommand{\smashproduct}{\wedge}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\proves}{\vdash}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Finite Types as Discrete (0-)Groupoids}

@@

%%%%%%%%%%%%%%%%%
\subsection{Types as Groupoids}

@@

\begin{definition}[Path]
A \emph{path} $p : x \path y$ between two points $x$ and $y$ in some space
$A$ is a directed edge $p$ from $x$ to $y$. The notation $\transport{p}{x}$
denotes the action of \emph{transporting} the point $x$ to the point~$y$
along $p$.
\end{definition}

\begin{definition}[Disjoint union of paths]
Given two sets of paths $P$ and $Q$, the disjoint union $P \uplus^p Q$ is
defined as:
\[ 
  \{ \leftv{p} : \leftv{x} \path \leftv{y} ~|~ (p : x \path y) \in P \}
  \union
  \{ \rightv{q} : \leftv{x} \path \leftv{y} ~|~ (q : x \path y) \in Q \}
\]
\end{definition}
We will sometimes write this as $\leftv{P} \union \rightv{Q}$.

\begin{definition}[Product of paths]
Given two sets of paths $P$ and $Q$, the product $P \times^p Q$ is
defined as:
\[ 
\{ (p,q) : (x,z) \path (y,w) ~|~ (p : x \path y) \in P, (q : z \path w) \in Q \}
\]
\end{definition}
We will sometimes write this as $(P,Q)$.

\begin{definition}[Groupoid]
A \emph{groupoid} $G = (G_0, \mathit{paths}_{G_0})$ is a set $G_0$ together
with a collection of paths $x \path y$, between elements $x$ and $y$ of $G_0$
that satisfies the following conditions:
\begin{itemize}
\item for every $x \in G_0$, there is a path $\refl_x : x \path x$,
\item every path $p : x \path y$, there is an inverse path $! p : y \path x$, 
\item for every paths $p : x \path y$ and $q : y \path z$ there is a path $p
  \circ q : x \path z$, and
\item for every paths $p : x \path y$, $q : y \path z$, and $r : z \path w$,
  we have:
  \begin{itemize}
  \item $\refl_x \circ p = p$;
  \item $p \circ \refl_y = p$;
  \item $(p \circ q) \circ r = (p \circ (q \circ r))$;
  \item $p \,\circ\, !p = \refl_x$;
  \item $!p \,\circ\, p = \refl_y$.
  \end{itemize}
\end{itemize}
\end{definition}
When introducing a groupoid, we only explicitly mention a collection of
non-trivial ``generating'' paths. The equivalence closure under $\refl$, $!$,
and $\circ$ with the appropriate conditions is implicitly
assumed.\footnote{For higher groupoids, the five equivalences in the last
  condition will be witnessed by paths between paths and not by the
  extensional equivalence relation $=$.}

\begin{definition}[Pointed groupoid]
A pointed groupoid $G_{\bullet} = (G_0, \bot_{G_0}, \mathit{paths}_{G_0})$ is
a groupoid $(G_0, \mathit{paths}_{G_0})$ with a distinguished element
$\bot_{G_0} \in G_0$.
\end{definition}

\begin{definition}[Wedge sum]
The \emph{wedge sum} of two pointed groupoids $G_\bullet$ and $H_\bullet$,
written $G_\bullet \wedgesum H_\bullet$ is defined as follows:
\[\begin{array}{l}
(G_0, \bot_{G_0}, \mathit{paths}_{G_0}) \wedgesum 
(H_0, \bot_{H_0}, \mathit{paths}_{H_0}) = \\
((G_0 \uplus H_0) \union \{ \bot \},
 \bot,
 (\mathit{paths}_{G_0} \uplus^p \mathit{paths}_{H_0}) \union
 \{ L: \leftv{\bot_{G_0}} \path \bot, 
    R: \rightv{\bot_{H_0}} \path \bot \})
\end{array}\]
where $\uplus$ is the disjoint union operation on sets and $\bot$ is a new
distinguished element that is related by new paths to the distinguished
elements of $G_\bullet$ and $H_\bullet$.
\end{definition}

\begin{definition}[Smash product]
The \emph{smash product} of two pointed groupoids $G_\bullet$ and
$H_\bullet$, written $G_\bullet \smashproduct H_\bullet$ is defined as
follows:
\[\begin{array}{l}
(G_0, \bot_{G_0}, \mathit{paths}_{G_0}) \smashproduct
(H_0, \bot_{H_0}, \mathit{paths}_{H_0}) = \\
((G_0 \times H_0) \union \{ \bot \},
 \bot,
 (\mathit{paths}_{G_0} \times^p \mathit{paths}_{H_0}) \union
 \{ F_h: (\bot_{G_0},h) \path \bot \} \union 
 \{ S_g: (g,\bot_{H_0}) \path \bot \} 
\end{array}\]
where $\times$ is the cartesian product operation on sets and $\bot$ is a new
distinguished element that is related by new paths to every pair of elements
that includes either $\bot_{G_0}$ or $\bot_{H_0}$. 
\end{definition}

\begin{definition}[Denotation of types]
Each type $b$ is mapped to a \emph{pointed groupoid} $(\dt{b}, \bot_b,
\mathit{paths}_b)$ as follows:
\[\begin{array}{rcl@{\qquad}l}
\dt{0} &=& (\{ \bot_0 \}, \bot_0, \{ \mathit{loop}_0 : \bot_0 \path \bot_0 \}) &
           (\mbox{circle}) \\
\dt{1} &=& (\{ \bullet, \bot_1 \}, \bot_1, 
           \{ \mathit{loop}_1 : \bot_1 \path \bot_1 \}) &\\
\dt{(b_1 + b_2)} &=& \dt{b_1} \wedgesum \dt{b_2} & 
           (\mbox{wedge~sum}) \\
\dt{(b_1 * b_2)} &=& \dt{b_1} \smashproduct \dt{b_2} & 
           (\mbox{smash~product})
\end{array}\]
\end{definition}

We now check that the wedge sum of any of our pointed groupoids $G_\bullet$
with $\dt{0}$ is ``essentially'' $G_\bullet$ itself, i.e., that $\dt{0}$ is
the additive unit in our space of pointed groupoids. Consider:
\[\begin{array}{l}
G_\bullet \wedgesum \dt{0} = \\
(G_0, \bot_{G_0}, \mathit{paths}_{G_0}) \wedgesum 
  (\{ \bot_0 \}, \bot_0, \{ \mathit{loop}_0 : \bot_0 \path \bot_0 \}) = \\
((\leftv{G_0} \union \{ \rightv{\bot_0}, \bot \}, 
 \bot,
 (\leftv{\mathit{paths}_{G_0}} \union \{ \rightv{\mathit{loop}^n_0} , 
 \leftv{\bot_{G_0}} \path \bot, \rightv{\bot_0} \path \bot \})
\end{array}\]
One can see that this space consists of $\leftv{G_0}$ with whatever structure
is on it and another connected component consisting of $\leftv{\bot_{G_0}}$,
$\rightv{\bot_0}$, and the new $\bot$ for the entire space. This cluster has
cardinality 0 and hence the new space is ``equivalent'' to the original
$G\bullet$ as we have essentially only changed the representation of 0.

>>>>

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage[sans]{dsfont}
\usepackage{color, colortbl}
\usepackage{courier}
\usepackage{graphicx}

@@

\newcommand{\union}{\cup}
\newcommand{\transport}[2]{\overline{#1}(#2)}
\newtheorem{definition}{Definition}
\newcommand{\refl}{\mathit{refl}}
\renewcommand{\path}{\leadsto}
\newcommand{\paths}[1]{\mathds{P}(#1)}
%% \renewcommand{\circ}{\circlearrowleft}
\newcommand{\alt}{~|~}
\newcommand{\dt}[1]{\llbracket #1 \rrbracket}
\newcommand{\leftv}[1]{\textsf{left}~#1}
\newcommand{\rightv}[1]{\textsf{right}~#1}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\wedgesum}{\vee}
\newcommand{\smashproduct}{\wedge}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\proves}{\vdash}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Category of Pointed Groupoids}

%%%%%%%%%%%%%%%%%
\subsection{Paths}

\begin{definition}[Path]
A \emph{path} $p : x \path y$ between two points $x$ and $y$ in some space
$A$ is a directed edge $p$ from $x$ to $y$. We denote the set of paths of the
space $A$ as $\paths{A}$.
\end{definition}

\begin{definition}[Disjoint union of paths]
Given $\paths{A}$ and $\paths{B}$ in spaces $A$ and $B$ respectively, we
construct $\paths{A} \uplus^p \paths{B}$ in space $A \uplus B$, the disjoint
union of $A$ and $B$, as follows:
\[ 
  \{ \leftv{p} : \leftv{x} \path \leftv{y} ~|~ 
    (p : x \path y) \in \paths{A} \}
  \union
  \{ \rightv{q} : \rightv{x} \path \rightv{y} ~|~ 
    (q : x \path y) \in \paths{B} \}
\]
\end{definition}
We will sometimes write this as $\leftv{(\paths{A})} \union
\rightv{(\paths{B})}$.

\begin{definition}[Product of paths]
Given $\paths{A}$ and $\paths{B}$ in spaces $A$ and $B$
respectively, we construct $\paths{A} \times^p \paths{B}$ in
space $A \times B$, the cartesian product of $A$ and $B$, as follows: 
\[ 
\{ (p,q) : (x,z) \path (y,w) ~|~ (p : x \path y) \in \paths{A}, 
   (q : z \path w) \in \paths{B} \}
\]
\end{definition}
We will sometimes write this as $(\paths{A},\paths{B})$.

%%%%%%%%%%%%%%%%%
\subsection{Pointed Groupoids}

\begin{definition}[Groupoid]
A \emph{groupoid} $G = (G_0, \paths{G_0})$ is a set $G_0$ together
with a collection of paths $x \path y$, between elements $x$ and $y$ of $G_0$
that satisfies the following conditions:
\begin{itemize}
\item for every $x \in G_0$, there is a path $\refl_x : x \path x$,
\item every path $p : x \path y$, there is an inverse path $! p : y \path x$, 
\item for every paths $p : x \path y$ and $q : y \path z$ there is a path $p
  \circ q : x \path z$, and
\item for every paths $p : x \path y$, $q : y \path z$, and $r : z \path w$,
  we have:
  \begin{itemize}
  \item $\refl_x \circ p = p$;
  \item $p \circ \refl_y = p$;
  \item $(p \circ q) \circ r = (p \circ (q \circ r))$;
  \item $p \,\circ\, !p = \refl_x$;
  \item $!p \,\circ\, p = \refl_y$.
  \end{itemize}
\end{itemize}
\end{definition}
When introducing a groupoid, we only explicitly mention a collection of
non-trivial ``generating'' paths. The equivalence closure under $\refl$, $!$,
and $\circ$ with the appropriate conditions is implicitly
assumed.\footnote{For higher groupoids, the five equivalences in the last
  condition will be witnessed by paths between paths and not by the
  extensional equivalence relation $=$.}

\begin{definition}[Pointed groupoid]
A pointed groupoid $G_{\bullet} = (G_0, \bot_{G_0}, \paths{G_0})$ is
a groupoid $(G_0, \paths{G_0})$ with a distinguished element
$\bot_{G_0} \in G_0$.
\end{definition}

\begin{definition}[Wedge sum]
The \emph{wedge sum} of two pointed groupoids $G_\bullet$ and $H_\bullet$,
written $G_\bullet \wedgesum H_\bullet$ is defined as follows:
\[\begin{array}{l}
(G_0, \bot_{G_0}, \paths{G_0}) \wedgesum 
(H_0, \bot_{H_0}, \paths{H_0}) = \\
((G_0 \uplus H_0) \union \{ \bot \},
 \bot,
 (\paths{G_0} \uplus^p \paths{H_0}) \union
 \{ L: \leftv{\bot_{G_0}} \path \bot, 
    R: \rightv{\bot_{H_0}} \path \bot \})
\end{array}\]
where $\uplus$ is the disjoint union operation on sets and $\bot$ is a new
distinguished element that is related by new paths to the distinguished
elements of $G_\bullet$ and $H_\bullet$.
\end{definition}

\begin{definition}[Smash product]
The \emph{smash product} of two pointed groupoids $G_\bullet$ and
$H_\bullet$, written $G_\bullet \smashproduct H_\bullet$ is defined as
follows:
\[\begin{array}{l}
(G_0, \bot_{G_0}, \paths{G_0}) \smashproduct
(H_0, \bot_{H_0}, \paths{H_0}) = \\
((G_0 \times H_0) \union \{ \bot \},
 \bot,
 (\paths{G_0} \times^p \paths{H_0}) \union
 \{ F_h: (\bot_{G_0},h) \path \bot \} \union 
 \{ S_g: (g,\bot_{H_0}) \path \bot \} 
\end{array}\]
where $\times$ is the cartesian product operation on sets and $\bot$ is a new
distinguished element that is related by new paths to every pair of elements
that includes either $\bot_{G_0}$ or $\bot_{H_0}$. 
\end{definition}

\begin{definition}[Denotation of types]
Each type $b$ is mapped to a \emph{pointed groupoid} $(\dt{b}, \bot_b,
\paths{b})$ as follows:
\[\begin{array}{rcl@{\qquad}l}
\dt{0} &=& (\{ \bot_0 \}, \bot_0, \{ \mathit{loop}_0 : \bot_0 \path \bot_0 \}) &
           (\mbox{circle}) \\
\dt{1} &=& (\{ \bullet, \bot_1 \}, \bot_1, 
           \{ \mathit{loop}_1 : \bot_1 \path \bot_1 \}) &\\
\dt{(b_1 + b_2)} &=& \dt{b_1} \wedgesum \dt{b_2} & 
           (\mbox{wedge~sum}) \\
\dt{(b_1 * b_2)} &=& \dt{b_1} \smashproduct \dt{b_2} & 
           (\mbox{smash~product})
\end{array}\]
\end{definition}

We now check that the wedge sum of any of our pointed groupoids $G_\bullet$
with $\dt{0}$ is ``essentially'' $G_\bullet$ itself, i.e., that $\dt{0}$ is
the additive unit in our space of pointed groupoids. Consider:
\[\begin{array}{l}
G_\bullet \wedgesum \dt{0} = \\
(G_0, \bot_{G_0}, \paths{G_0}) \wedgesum 
  (\{ \bot_0 \}, \bot_0, \{ \mathit{loop}_0 : \bot_0 \path \bot_0 \}) = \\
((\leftv{G_0} \union \{ \rightv{\bot_0}, \bot \}, 
 \bot,
 (\leftv{\paths{G_0}} \union \{ \rightv{\mathit{loop}^n_0} , 
 \leftv{\bot_{G_0}} \path \bot, \rightv{\bot_0} \path \bot \})
\end{array}\]
One can see that this space consists of $\leftv{G_0}$ with whatever structure
is on it and another connected component consisting of $\leftv{\bot_{G_0}}$,
$\rightv{\bot_0}$, and the new $\bot$ for the entire space. This cluster has
cardinality 0 and hence the new space is ``equivalent'' to the original
$G\bullet$ as we have essentially only changed the representation of 0.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Finite Types as Pointed Groupoids}

@@

@@
